---
title: 'Stat 537: Homework 4'
author: "Brandon Fenton and Kenny Flagg"
date: "Due Friday, Feb 12 at end of day"
output: pdf_document
header-includes: \usepackage{float}
---

1. _For the ranked data, make a pairs plot of bagplots (using bagplot.pairs). Discuss and identify (by location and name of institution) any potential outliers clearly identified in the pair-wise plots._

```{r p1_a, echo=F, comment=NA, message=F, warning=F, fig.width=15, fig.height=12, out.width="\\linewidth", out.height="0.75\\linewidth", fig.pos="H", fig.align="center", cache=T}
#fig.cap="Pairwise bagplots",
require(aplpack)

cc2015<-read.csv("https://montana.box.com/shared/static/0tp7ooud5sux4wzpkrjgklpudrreu6jc.csv",header=T,na.strings="dnr")
cc2015_PhD<-cc2015[(cc2015$BASIC2015>14&cc2015$BASIC2015<18),]
cc2015_PhD$BASIC2015<-factor(cc2015_PhD$BASIC2015)
cc2015Ps<-cc2015_PhD[,c("NAME","BASIC2010","BASIC2015","FACNUM","HUM_RSD","OTHER_RSD","SOCSC_RSD","STEM_RSD","PDNFRSTAFF","S.ER.D","NONS.ER.D")]
cc2015Ps2<-na.omit(cc2015Ps)
cc2015Rs<-data.frame(cc2015Ps2[,1:3],sapply(cc2015Ps2[,-c(1:3)],rank))
cc2015Rs_2<-cc2015Rs[,c("HUM_RSD","OTHER_RSD","SOCSC_RSD","STEM_RSD","PDNFRSTAFF","S.ER.D","NONS.ER.D")]

#cc2015Ps2[cc2015Ps2$NAME=="Montana State University",]
#cc2015Rs[cc2015Rs$NAME=="Montana State University",]

invisible(bagplot.pairs(cc2015Rs_2))
```

UC Denver and Rockefeller University are potential outliers in __STEM\_RSD__ and __PDNFRSTAFF__.
```{r p1_b, echo=F, comment=NA, message=F, warning=F}
cc2015Rs[(cc2015Rs_2$STEM_RSD<100 & cc2015Rs_2$PDNFRSTAFF>200) |
           (cc2015Rs_2$STEM_RSD>160 & cc2015Rs_2$PDNFRSTAFF<30),
         c(1, 8, 9)]
```

CUNY Graduate School is a potential outlier in __STEM\_RSD__ and __S.ER.D__.
```{r p1_c, echo=F, comment=NA, message=F, warning=F}
cc2015Rs[(cc2015Rs_2$STEM_RSD>200 & cc2015Rs_2$S.ER.D<50), c(1, 8, 10)]
```

UC Denver is also potentially an outlier in __PDNFRSTAFF__ and __S.ER.D__.
```{r p1_d, echo=F, comment=NA, message=F, warning=F}
cc2015Rs[(cc2015Rs_2$PDNFRSTAFF<50 & cc2015Rs_2$S.ER.D>200), c(1, 9, 10)]
```

Denver is tied for the lowest rank in terms of PDNFRSTAFF; however none of these institutions appear to be outliers in any single variable so they would not have been discovered as possible outliers had we not examined the pairwise plots.


2. _Read the help file on the rank function. What is the default choice for dealing with ties? How does this show up in the plots of the data set?_

The __rank__ function allows several ways of dealing with ties via the __ties.method__ argument, but the default method is to assign the mean of the ranks the tied observations would hold to each of the tied observations.  If this method is used, rank-based plots involving tied values will contain gaps corresponding to the ranks which have not been assigned.  For example, __rank(c(5,6,6,6,7))__ would assign a rank of 3 to the middle three values, and a plot of ranked values would contain empty space in locations corresponding to ranks 2 and 4 for that variable.



3. _Make a plot that includes the correlations. Discuss the strength and directions of correlations. Do you expect to require many dimensions in the PCA to capture the information in these variables?_

```{r p3_a, echo=F, comment=NA, message=F, warning=F, out.width="\\linewidth", out.height="0.5\\linewidth", fig.width=14, fig.height=7, fig.align="center", fig.cap="All pairwise correlations", fig.pos="H"}
require(corrplot)
cor.mat <- cor(cc2015Rs_2)
corrplot.mixed(cor.mat, srt = -45)
```

All of the correlations are positive and most are moderate to strong. The weakest correlations are between __OTHER\_RSD__ and __STEM\_RSD__, __PDNFRSTAFF__, and __S.ER.D__. These are weak to moderate correlations of around 0.4. Given the strong associations between the variables we expect to need very few dimensions to capture the variation in the data, with most of the variation explained by the first principal component.

\pagebreak
4. _Do a PCA of the correlation matrix based on the ranked data using prcomp._

a. _Make a nice plot of the eigenvalues (not a bar chart) and discuss the choice of the number of components to use based on the elbow, cumulative percentage of variance, the percentage of variance for each additional component, the latent root criterion and the “simulation” based recommendation discussed in EH. How many would you use here?_

```{r p4_a, echo=F, comment=NA, message=F, warning=F}
require(pander)

pcs <- prcomp(cor.mat, scale=T)
evals <- pcs$sdev^2
ev.total <- sum(evals)
ev.props <- evals/ev.total
ev.prop.cum <- cumsum(ev.props)
pander(data.frame(PC=1:7, Eigenvalue=signif(evals,3), Proportion=signif(ev.props,3), 
                  Cumulative =signif(ev.prop.cum,3)))
```
```{r p4_b, echo=F, comment=NA, message=F, warning=F, out.width="\\linewidth", out.height="0.6\\linewidth", fig.width=12, fig.height=6, fig.pos="H", fig.cap="Scree Plot"}
plot(pcs, type="lines", main = "Eigenvalues")
```
The elbow of the plot occurs at PC2, indicating that the principal components beyond the first two have relatively small individual contributions to the observed variation. The first two PCs explain 80.9\% of the variance; the third PC brings the cumulative total to 95.0\% and the first four PCs describe 99.1\% of the variance, meaning that nearly all of the variation can be described in four dimensions. The individual percentages of variance for the first four PCs are, respectively, 55.5\%, 25.4\%, 14.1\%, and 4.09\%. Based on the individual contributions we might use only the first three pricipal components because the fourth has a relatively unsubstantial contribution. By the latent root criterion, the first two PCs would be used as they are the only ones that contribute more than expected; PC3 comes close to the critical variance of 1 and would be included by Jolliffe's threshold of 0.7. All things considered, we would use the first three principal components because they contribute 95\% of the total variation, leaving very little variation in any additional PC.

\pagebreak
b. _Use the eigenvector coefficients to write out interpretations of the first 3 PCs._

```{r p4_c, echo=F, comment=NA, message=F, warning=F}
pander(pcs)
```

PC1 is a weighted average of __STEM\_RSD__, __PDNFRSTAFF__, and __S.ER.D__ contrasted with __OTHER\_RSD__. PC2 is an average of __HUM\_RSD__ and __SOCSC\_RSD__, and PC3 is __NONS.ER.D__.

5. _Make a biplot. First, discuss how good of a picture of the data set this is based on the percentage of variation these two PCs contain (is it a “good picture” or is it leaving a lot out?). Discuss the observation that is labelled at 181 based on the biplot. What does the biplot suggest about its values on the two PCs and each of the 7 variables? You should use words like “about average” or discuss much above or below average it is on each in this interpretation. This is about what you are getting from the plot and not necessarily what is in the ranked variables. One would hope that they agree, but that is not always the case._
```{r p5_a, echo=F, comment=NA, message=F, warning=F, out.width="\\linewidth", out.height="0.3\\linewidth", fig.width=12, fig.height=4}
biplot(pcs)
```

6. _Plot the first two PC scores with the BASIC2015 group information and highlight MSU in this plot. Discuss its location relative to the other schools – which group does it seem to belong to?_
```{r p6_a, echo=F, comment=NA, message=F, warning=F, out.width="0.8\\linewidth", out.height="0.8\\linewidth", fig.width=6, fig.height=6}
#something weird going on here
pc.scores <- as.matrix(cc2015Rs_2) %*% predict(pcs)[,1:2]
score.col <- as.numeric(cc2015Rs$BASIC2015) + 3
score.pch <- score.col
score.col[which(cc2015Rs$NAME=="Montana State University")] <- 2
score.pch[which(cc2015Rs$NAME=="Montana State University")] <- 19
plot(pc.scores, col=score.col, pch=score.pch)
legend(x=max(pc.scores[,1])*0.7, y=max(pc.scores[,2])*0.97,legend=c(unique(cc2015Rs$BASIC2015),"MSU"), 
       col = unique(score.col2), pch=unique(score.pch2))

```


7. _The actual Carnegie scale groups are developed also using information from a PCA of 3 per capita variables for staff and the two types of expenditures divided by number of faculty. Repeat the PCA of the correlation matrix of the ranked data for these three variables, make a biplot, and interpret the first two PCs._
```{r p7_a, echo=F, comment=NA, message=F, warning=F, out.width="\\linewidth", out.height="0.3\\linewidth", fig.width=12, fig.height=4}
cc2015percap<-cc2015Ps2[,c("PDNFRSTAFF","S.ER.D","NONS.ER.D")]/cc2015Ps2$FACNUM
cc2015percap_rank<-data.frame(sapply(cc2015percap,rank))
pcpcs <- prcomp(cor(cc2015percap_rank), scale=T)
# bagplot.pairs(cc2015percap_rank)
# pairs.panels(cc2015percap_rank)
percevals <- pcpcs$sdev^2
percev.total <- sum(percevals)
percev.props <- percevals/percev.total
percev.prop.cum <- cumsum(percev.props)
pander(data.frame(PC=1:3, Eigenvalue = signif(percevals,3), Proportion = signif(percev.props,3), 
                  Cumulative = signif(percev.prop.cum,3)))
```

```{r p7_b, echo=F, comment=NA, message=F, warning=F, out.width="\\linewidth", out.height="0.6\\linewidth", fig.width=12, fig.height=6}
plot(pcpcs, type="lines", main = "Eigenvalues")
```

```{r p7_c, echo=F, comment=NA, message=F, warning=F, out.width="\\linewidth", out.height="0.3\\linewidth", fig.width=12, fig.height=4}
biplot(pcpcs)
```

8. _Make a plot of the first PC from the “aggregated” and ranked data and the first PCA from this per capita ranked data with the aggegrated data PC score on the x-axis. Orient the plots so that higher is better, add information from BASIC2015 to the plot and indicate/flag MSU in some fashion._
```{r p8_a, echo=F, comment=NA, message=F, warning=F, out.width="0.8\\linewidth", out.height="0.8\\linewidth", fig.width=6, fig.height=6}
pcpc.scores <- as.matrix(cc2015percap_rank) %*% predict(pcpcs)[,1]
score.col2 <- as.numeric(cc2015Rs$BASIC2015) + 3
score.pch2 <- score.col
score.col2[which(cc2015Rs$NAME=="Montana State University")] <- 2
score.pch2[which(cc2015Rs$NAME=="Montana State University")] <- 19
plot(x=pc.scores[,1], y=pcpc.scores, col = score.col2, pch=score.pch2)

legend(x=max(pc.scores[,1])*0.7, y=-100,legend=c(unique(cc2015Rs$BASIC2015),"MSU"), 
       col = unique(score.col2),  pch=unique(score.pch2))
```

9. _Discuss MSU’s location on these two dimensions and the separation among the three Carnegie classifications._
```{r p9_a, echo=F, comment=NA, message=F, warning=F, out.width="\\linewidth", out.height="0.3\\linewidth", fig.width=12, fig.height=4}
```

## R Code Appendix:
Problem 1:
```{r a1, ref.label='p1_a', eval=F}
```
```{r b1, ref.label='p1_b', eval=F}
```
```{r b1, ref.label='p1_c', eval=F}
```
```{r b1, ref.label='p1_d', eval=F}
```

Problem 3:
```{r a3, ref.label='p3_a', eval=F}
```

Problem 4:
```{r a4, ref.label='p4_a', eval=F}
```
```{r b4, ref.label='p4_b', eval=F}
```
```{r c4, ref.label='p4_c', eval=F}
```

Problem 5:
```{r a5, ref.label='p5_a', eval=F}
```

```{r a5b, ref.label='p5_b', eval=F}
```

Problem 6:
```{r a6, ref.label='p6_a', eval=F}
```

Problem 7:
```{r a7, ref.label='p7_a', eval=F}
```

```{r a7b, ref.label='p7_b', eval=F}
```

```{r a7c, ref.label='p7_c', eval=F}
```

Problem 8:
```{r a8, ref.label='p8_a', eval=F}
```

Problem 9:
```{r a9, ref.label='p9_a', eval=F}
```

### About This Markdown File

  * File creation date: `r Sys.Date()`
  * `r R.version.string`
  * R version (short form): `r getRversion()`
  * Additional session information
  
```{r echo=FALSE}
sessionInfo()  # could use devtools::session_info() if you prefer that
```
